---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 3 - Due date 02/10/23"
author: "Kexin(Kassie) Huang"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(Kendall)
```

##Trend Component

### Q1
Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code form A2, but I want all three plots on the same window this time. (Hint: use par() function)
```{r}
#import data
library(readxl)
energy_data <- read_excel(path='./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx',sheet=1,skip=12,col_names=FALSE)
colnames(energy_data) <- read_excel(path='./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx',sheet=1, skip=10, n_max=1, col_names=FALSE)
energy_data <- as.data.frame(energy_data[,4:6])
ts_energy_data <- ts(energy_data,start = c(1973,1), frequency = 12)

#index
datanames <- colnames(energy_data)
ncols <- ncol(energy_data)

#Time series plot
library(ggplot2)
library(ggfortify)
library(patchwork)

plots_ts <- list()
for(i in 1:ncols){
  p <-  autoplot(ts_energy_data[,i])+
    labs(title = datanames[i],y='Trillion Btu')+
    theme(text = element_text(size = 6))
  plots_ts[[i]] <- p
}
wrap_plots(plots_ts, ncol = 3)

#ACF&PACF
par(mfrow=c(1,3))
for(i in 1:ncols){
  Acf(ts_energy_data[,i],lag.max=40,main=datanames[i],cex.main=0.5,ylim=c(-1,1),ylab='Acf')
}
for(i in 1:ncols){
  Pacf(ts_energy_data[,i],lag.max=40,main=datanames[i],cex.main=0.5,ylim=c(-1,1),ylab='Pacf')
}
```


### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

Answer:
Yes, all of them have a trend. Both the total biomass energy production and the total renewable energy production have an increase trend, and the hydroelectric power consumption has a slight decrease trend.

*what do you mean by "what kind of trend"? deterministic vs stochastic?*

### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.
```{r}

t <- 1:nrow(energy_data)


#Biomass Energy Production
linear_trend_biomass <- lm(energy_data$`Total Biomass Energy Production`~t)
summary(linear_trend_biomass)
beta0_biomass <- linear_trend_biomass$coefficients[1]
beta1_biomass <- linear_trend_biomass$coefficients[2]

#Renewable Energy Production
linear_trend_renewable <- lm(energy_data$`Total Renewable Energy Production`~t)
summary(linear_trend_renewable)
beta0_renewable <- linear_trend_renewable$coefficients[1]
beta1_renewable <- linear_trend_renewable$coefficients[2]

#Hydroelectric Power Consumption
linear_trend_hydro <- lm(energy_data$`Hydroelectric Power Consumption`~t)
summary(linear_trend_hydro)
beta0_hydro <- linear_trend_hydro$coefficients[1]
beta1_hydro <- linear_trend_hydro$coefficients[2]

```
 
### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?
```{r}
#create an empty list to store plots
plots2_ts <- list()

#plot the de-trended series and compare with the original one
detrend_biomass <- energy_data[,1]-(beta0_biomass + beta1_biomass*t)
ts_detrend_biomass <- ts(detrend_biomass,start = c(1973.1),frequency=12)
p <- autoplot(ts_detrend_biomass)+
  labs(title=datanames[1],y='Trillion btu')+
  theme(text=element_text(size=6))
plots2_ts[[1]] <- p
plots2_ts[[1]] + plots_ts[1]


```

### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?
```{r}
par(mfrow=c(1,2))
Acf(ts_detrend_biomass,lag.max = 40,ylim=c(-1,1),cex.main=0.6)
Pacf(ts_detrend_biomass,lag.max = 40,ylim=c(-1,1),cex.main=0.6)

```

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.

### Q7
Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

### Q8
Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?


