---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
author: "Kexin(Kassie) Huang"
subtitle: "Assignment 8 - Due date 03/27/23"
output:
  pdf_document: default
  html_document:
    df_print: paged
geometry: margin=2.54cm
editor_options:
  chunk_output_type: console
---

## Directions

You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. And to do so you will need to fork our repository and link it to your RStudio.

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. Then you will start working through the assignment by **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A08_Sp22.Rmd"). Submit this pdf using Sakai.

## Set up

Some packages needed for this assignment: `forecast`,`tseries`,`smooth`. Do not forget to load them before running your script, since they are NOT default packages.

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(smooth)
library(ggplot2)
library(tidyverse)
library(lubridate)

```

## Importing and processing the data set

Consider the data from the file "inflowtimeseries.txt". The data corresponds to the monthly inflow in $m^{3}/s$ for some hydro power plants in Brazil. You will only use the last column of the data set which represents one hydro plant in the Amazon river basin. The data span the period from January 1931 to August 2011 and is provided by the Brazilian ISO.

For all parts of the assignment prepare the data set such that the model consider only the data from January 2000 up to December 2009. Leave the year 2010 of data (January 2010 to December 2010) for the out-of-sample analysis. Do **NOT** use data fro 2010 and 2011 for model fitting. You will only use it to compute forecast accuracy of your model.

## Part I: Preparing the data sets

### Q1

Read the file into a data frame. Prepare your time series data vector such that observations start in January 2000 and end in December 2009. Make you sure you specify the **start=** and **frequency=** arguments. Plot the time series over time, ACF and PACF.

```{r}

#read data and prepare the ts
df_data <- read.table("./Data/inflowtimeseries.txt", header = FALSE)
df_inflow_all <- data.frame(cbind(paste(df_data$V1,df_data$V2),df_data$V17)) %>%
  rename( Date = X1, Inflow = X2) %>%
  mutate( Date = my(Date), Inflow = as.numeric(Inflow))
df_inflow_09 <- df_inflow_all %>%
  filter(Date >= as.Date("2000-01-01") & Date <= as.Date("2009-12-31"))

ts_inflow_all <- ts(df_inflow_all$Inflow,start = c(1931,1),frequency = 12)

#extract the observations from 2000-2009
ts_inflow_09 <- window(ts_inflow_all, start = c(2000, 1), end = c(2009, 12))

#plot the time series, ACF and PACF
par(mar=c(3,3,3,0));par(mfrow=c(1,3))
ts.plot(ts_inflow_09,main="Time series plot")
Acf(ts_inflow_09,lag.max=40,ylim=c(-1,1),main="ACF plot")
Pacf(ts_inflow_09,lag.max=40,ylim=c(-1,1),main="PACF plot")

```

### Q2

Using the $decompose()$ or $stl()$ and the $seasadj()$ functions create a series without the seasonal component, i.e., a deseasonalized inflow series. Plot the deseasonalized series and original series together using ggplot, make sure your plot includes a legend. Plot ACF and PACF for the deaseasonalized series. Compare with the plots obtained in Q1.

```{r}

#decompose and deseason
decompose_inflow_all <- stl(ts_inflow_all, s.window = "periodic")
deseason_inflow_all <- seasadj(decompose_inflow_all)
deseason_inflow_09 <- window(deseason_inflow_all, start = c(2000, 1), end = c(2009, 12))

#plot the original and deseasoned series
combined_09 <- data.frame(Date = df_inflow_09$Date,
                           Original = df_inflow_09$Inflow,
                           Deseasonal = deseason_inflow_09)

ggplot(combined_09, aes(x = Date)) +
  geom_line(aes(y = Original, color = "Original"), size = 1) +
  geom_line(aes(y = Deseasonal, color = "Deseasonal"), size = 1) +
  scale_color_manual(values = c("Original" = "black", "Deseasonal" = "blue")) +
  labs(x = "Date", y = "Inflow",title = "Deseasonal and Original Series", color = "Legend") 

#ACF and PACF plot
par(mar=c(3,3,3,0));par(mfrow=c(2,3))
ts.plot(deseason_inflow_09,main="Time series plot (deseasonal)")
Acf(deseason_inflow_09,lag.max=40,ylim=c(-1,1),main="ACF plot")
Pacf(deseason_inflow_09,lag.max=40,ylim=c(-1,1),main="PACF plot")

ts.plot(ts_inflow_09,main="Time series plot (original)")
Acf(ts_inflow_09,lag.max=40,ylim=c(-1,1),main="ACF plot")
Pacf(ts_inflow_09,lag.max=40,ylim=c(-1,1),main="PACF plot")
```

> Answer:
Compared to the original data, the seasonal pattern is less pronounced, as shown in all three plots: The regular wave pattern in the original time series plot no longer exists in the deseasonal time series plot. Also, the magnitude of the "wave" in the ACF plot decreased after deseasoning. In the PACF plot, the number of spikes exceeding the significance interval decreases.

## Part II: Forecasting with ARIMA models and its variations

### Q3

Fit a non-seasonal ARIMA$(p,d,q)$ model using the auto.arima() function to the non-seasonal data. Forecast 12 months ahead of time using the $forecast()$ function. Plot your forecasting results and further include on the plot the last year of non-seasonal data to compare with forecasted values (similar to the plot on the lesson file for M10).

```{r}
#forecast
par(mfrow=c(1,1))
deseason_09_arima <- auto.arima(deseason_inflow_09, seasonal = FALSE)
forecast_deaseason_09_arima <- forecast(deseason_09_arima, h = 12)
plot(forecast_deaseason_09_arima)
                             
#extract the actual non-seasonal data
deseason_inflow_10 <- window(deseason_inflow_all,start=c(2010,1),end=c(2010,12))

#combine the forecast result with the actual data
autoplot(window(deseason_inflow_all,start=c(2000,1),end=c(2010,12)),
         xlab="Year",ylab="Inflow")+
  autolayer(forecast_deaseason_09_arima,series="Non-seasonal ARIMA",PI=FALSE)+
  autolayer(deseason_inflow_10, series="Actual non-seasonal data")

```

> Answer:
The non-seasonal ARIMA produced a series of constant, over-estimated forecasted values compared with the actual data.

### Q4

Put the seasonality back on your forecasted values and compare with the original seasonal data values. $Hint:$ One way to do it is by summing the last year of the seasonal component from your decompose object to the forecasted series. 


```{r}

#generate the seasonal component
season_only_09 <- window(decompose_inflow_all$time.series[,"seasonal"],start=c(2000,1),end=c(2009,12)) 
forecast_season_09 <- forecast(season_only_09, h=12)
plot(forecast_season_09)

#add the seasonal component to previous ARIMA forecast
forescast_deaseason_09_arima_added <- forecast_deaseason_09_arima$mean + forecast_season_09$mean 

#extract the original data
original_inflow_10 <- window(ts_inflow_all,start=c(2010,1),end=c(2010,12))

#combine the new series with the original data
autoplot(window(ts_inflow_all,start=c(2000,1),end=c(2010,12)),
         xlab="Year",ylab="Inflow")+
  autolayer(forescast_deaseason_09_arima_added,series="Non-seasonal ARIMA, seasons added")+
  autolayer(original_inflow_10, series="Actual seasonal data")

```

### Q5

Repeat Q3 for the original data, but now fit a seasonal ARIMA$(p,d,q)x(P,D,Q)_ {12}$ also using the auto.arima().

```{r}
#forecast
original_09_arima <- auto.arima(ts_inflow_09)
forecast_original_09_arima <- forecast(original_09_arima, h = 12)
plot(forecast_original_09_arima)

#combine the forecast result with the actual non-seasonal data
autoplot(window(ts_inflow_all,start=c(2000,1),end=c(2010,12)),
         xlab="Year",ylab="Inflow")+
  autolayer(forecast_original_09_arima,series="Seasonal ARIMA",PI=FALSE)+
  autolayer(original_inflow_10, series="Actual seasonal data")
```

### Q6

Compare the plots from Q4 and Q5 using the autoplot() function.

```{r}
autoplot(window(ts_inflow_all,start=c(2000,1),end=c(2010,12)),
         xlab="Year",ylab="Inflow")+
  autolayer(forecast_original_09_arima,series="Seasonal ARIMA",PI=FALSE)+
  autolayer(forescast_deaseason_09_arima_added,series="Non-seasonal ARIMA, seasons added")+
  autolayer(original_inflow_10, series="Actual seasonal data")
```

The SARIMA model performs better forecast than using manually de-seasoning and fitting a non-seasonal ARIMA: the forecast from SARIMA is always closer to the actual seasonal data than the other.

## Part III: Forecasting with Other Models

### Q7

Fit an exponential smooth model to the original time series using the function $ses()$  from package `forecast`. Note that this function automatically do the forecast. Do not forget to set the arguments: silent=FALSE and holdout=FALSE, so that the plot is produced and the forecast is for the year of 2010.

```{r}
original_09_es <-  es(ts_inflow_09,silent=FALSE,holdout=FALSE)
forecast_original_09_es <- forecast(original_09_es,h=12)
plot(forecast_original_09_es)

```


## Part IV: Checking Forecast Accuracy

### Q8

Make one plot with the complete original seasonal historical data (Jan 2000 to Dec 2010). Now add the forecasts from each of the developed models in parts Q4, Q5, Q7 and Q8. You can do it using the autoplot() combined with autolayer(). If everything is correct in terms of time line, the forecasted lines should appear only in the final year. If you decide to use ggplot() you will need to create a data frame with all the series will need to plot. Remember to use a different color for each model and add a legend in the end to tell which forecast lines corresponds to each model.


```{r}
#plot together
autoplot(window(ts_inflow_all,start=c(2000,1),end=c(2010,12)),
         xlab="Year",ylab="Inflow")+
  autolayer(forecast_original_09_arima,series="Seasonal ARIMA",PI=FALSE)+
  autolayer(forescast_deaseason_09_arima_added,series="Non-seasonal ARIMA, seasons added")+
  autolayer(forecast_original_09_es,series="Exponential smooth model")+
  autolayer(original_inflow_10, series="Actual seasonal data")

```

### Q9

From the plot in Q9 which model or model(s) are leading to the better forecasts? Explain your answer. Hint: Think about which models are doing a better job forecasting the high and low inflow months for example.

SARIMA works the best, as it appears to be the closest to the actual data, especially at the seasonal peak and trough of the forecasted year.

### Q10

Now compute the following forecast metrics we learned in class: RMSE and MAPE,  for all the models you plotted in part Q9. You can do this by hand since your have forecasted and observed values for the year of 2010. Or you can use R function $accuracy()$ from package "forecast" to do it. Build and a table with the results and highlight the model with the lowest MAPE. Does the lowest MAPE corresponds match your answer for part Q10?

```{r}
ARIMA_scores <- accuracy(forescast_deaseason_09_arima_added,x = original_inflow_10)
SARIMA_scores <- accuracy(forecast_original_09_arima$mean,x = original_inflow_10)
ES_scores <- accuracy(forecast_original_09_es$mean, x=original_inflow_10)

seas_scores <- as.data.frame(rbind(ARIMA_scores,SARIMA_scores,ES_scores))
row.names(seas_scores) <- c("ARIMA", "SARIMA","ES")

print(seas_scores)
```

Yes. SARIMA seems to have the lowest MAPE, indicating it's the best model.
