---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2023"
subtitle: "Assignment 3 - Due date 02/10/23"
author: "Kexin(Kassie) Huang"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the file open on your local machine the first thing you will do is rename the file such that it includes your first and last name (e.g., "LuanaLima_TSA_A02_Sp23.Rmd"). Then change "Student Name" on line 4 with your name. 

Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the December 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only. 

R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here
library(forecast)
library(tseries)
library(Kendall)
```

##Trend Component

### Q1
Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code form A2, but I want all three plots on the same window this time. (Hint: use par() function)

Answer:

```{r}
#import data
library(readxl)
energy_data <- read_excel(path='./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx',sheet=1,skip=12,col_names=FALSE)
colnames(energy_data) <- read_excel(path='./Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx',sheet=1, skip=10, n_max=1, col_names=FALSE)
energy_data <- as.data.frame(energy_data[,4:6])
ts_energy_data <- ts(energy_data,start = c(1973,1), frequency = 12)

#index
datanames <- colnames(energy_data)
ncols <- ncol(energy_data)

#Time series plot
library(ggplot2)
library(ggfortify)
library(patchwork)

plots_ts <- list()
for(i in 1:ncols){
  p <-  autoplot(ts_energy_data[,i])+
    labs(title = datanames[i],y='Trillion Btu')+
    theme(text = element_text(size = 6))
  plots_ts[[i]] <- p
}
wrap_plots(plots_ts, ncol = 3)

#ACF&PACF
par(mfrow=c(1,3))
for(i in 1:ncols){
  Acf(ts_energy_data[,i],lag.max=40,main=datanames[i],cex.main=0.5,ylim=c(-1,1),ylab='Acf')
}
for(i in 1:ncols){
  Pacf(ts_energy_data[,i],lag.max=40,main=datanames[i],cex.main=0.5,ylim=c(-1,1),ylab='Pacf')
}
```


### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

Answer:

Yes, all of them have a trend. Both the total biomass energy production and the total renewable energy production have an increasing trend, and the hydroelectric power consumption has a slightly decreasing trend.


### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

Answer:
```{r}

nobs <- nrow(energy_data)
t <- 1:nobs


#Biomass Energy Production
linear_trend_biomass <- lm(energy_data$`Total Biomass Energy Production`~t)
summary(linear_trend_biomass)
beta0_biomass <- linear_trend_biomass$coefficients[1]
beta1_biomass <- linear_trend_biomass$coefficients[2]

#Renewable Energy Production
linear_trend_renewable <- lm(energy_data$`Total Renewable Energy Production`~t)
summary(linear_trend_renewable)
beta0_renewable <- linear_trend_renewable$coefficients[1]
beta1_renewable <- linear_trend_renewable$coefficients[2]

#Hydroelectric Power Consumption
linear_trend_hydro <- lm(energy_data$`Hydroelectric Power Consumption`~t)
summary(linear_trend_hydro)
beta0_hydro <- linear_trend_hydro$coefficients[1]
beta1_hydro <- linear_trend_hydro$coefficients[2]

```
 
Interpretation:
For biomass energy production, the coefficient of t is 0.48 with the p-value less than 0.001, and the intercept is 133.7. In other words, the biomass energy production is positively correlated with time, and the linear trend model can be written as:
Biomass energy production = 134 + 0.48*t

Similarly, the p-value of the remaining two models are both less than 0.001, which indicates significant correlation. The linear trend models can be written as:
Renewable energy production = 312 + 0.936*t
Hydroelectric power consumption = 260 - 0.0829*t 

### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

Answer:
```{r}
#create an empty list to store plots
plots2_ts <- list()

#plot the de-trended series and compare with the original one
detrend_biomass <- energy_data[,1]-(beta0_biomass + beta1_biomass*t)
ts_detrend_biomass <- ts(detrend_biomass,start = c(1973.1),frequency=12)
print(autoplot(ts_detrend_biomass)+
  labs(title=paste(datanames[1],"(Black: detrended)"),y='Trillion btu')+
  theme(text=element_text(size=6))+
  geom_line(aes(y=ts_energy_data[,1]),color='brown')+
  geom_hline(yintercept = 0, linetype="dashed")
  )

detrend_renewable <- energy_data[,2]-(beta0_renewable + beta1_renewable*t)
ts_detrend_renewable <- ts(detrend_renewable,start = c(1973.1),frequency=12)
print(autoplot(ts_detrend_renewable)+
  labs(title=paste(datanames[2],"(Black: detrended)"),y='Trillion btu')+
  theme(text=element_text(size=6))+
  geom_line(aes(y=ts_energy_data[,2]),color='green')+
  geom_hline(yintercept = 0, linetype="dashed")
)

detrend_hydro <- energy_data[,3]-(beta0_hydro + beta1_hydro*t)
ts_detrend_hydro <- ts(detrend_hydro,start = c(1973.1),frequency=12)
print(autoplot(ts_detrend_hydro)+
  labs(title=paste(datanames[3],"(Black: detrended)"),y='Trillion btu')+
  theme(text=element_text(size=6))+
  geom_line(aes(y=ts_energy_data[,3]),color='blue')+
  geom_hline(yintercept = 0, linetype="dashed"))
```

All of the plot changed after being detrended.
Firstly, the values of y-axis all decreased a lot, and the detrended lines all flucuate around 0.
Also, in terms of the shape of detrended lines, the first half of biomass's line seems to be "lifted up" a bit, making the overall trend more horizontal. Same for renewable energy production. Hydro power consumption changed in the opposite, but the overall trend also get more horizontal.

### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?
```{r}

#Biomass energy production
par(mfrow=c(1,2))
Acf(ts_energy_data[,1],lag.max=40,main=datanames[1],cex.main=0.3,ylim=c(-1,1))
Acf(ts_detrend_biomass,lag.max = 40, main=paste(datanames[1],"(detrended)"),ylim=c(-1,1),cex.main=0.3, )
Pacf(ts_energy_data[,1],lag.max=40,main=datanames[1],cex.main=0.3,ylim=c(-1,1))
Pacf(ts_detrend_biomass,lag.max = 40, main=paste(datanames[1],"(detrended)"),ylim=c(-1,1),cex.main=0.3)
```
After detrending, most of the values in the ACF plot decreased, except for the values around lag=12, 24, and 36. Likewise, in the PACF plot, only the values around lag=12, 24, and 36 increased.

```{r}
#Renewable energy production
par(mfrow=c(1,2))
Acf(ts_energy_data[,2],lag.max=40,main=datanames[2],cex.main=0.3,ylim=c(-1,1))
Acf(ts_detrend_renewable,lag.max = 40, main=paste(datanames[2],"(detrended)"),ylim=c(-1,1),cex.main=0.3, )
Pacf(ts_energy_data[,2],lag.max=40,main=datanames[2],cex.main=0.3,ylim=c(-1,1))
Pacf(ts_detrend_renewable,lag.max = 40, main=paste(datanames[2],"(detrended)"),ylim=c(-1,1),cex.main=0.3)
```
Most of the values in the ACF plot have decreased after the detrending, and the wave-like shape is more pronounced. On the other hand, the values in the PACF plot increased after detrending, indicating a stronger autocorrelation.

```{r}
#Hydro energy consumption
par(mfrow=c(1,2))
Acf(ts_energy_data[,3],lag.max=40,main=datanames[3],cex.main=0.3,ylim=c(-1,1))
Acf(ts_detrend_hydro,lag.max = 40, main=paste(datanames[3],"(detrended)"),ylim=c(-1,1),cex.main=0.3, )
Pacf(ts_energy_data[,3],lag.max=40,main=datanames[3],cex.main=0.3,ylim=c(-1,1))
Pacf(ts_detrend_hydro,lag.max = 40, main=paste(datanames[3],"(detrended)"),ylim=c(-1,1),cex.main=0.3)

```
The ACF plot shows a stronger, clearer seasonal component after detrending, with deeper troughs and narrower waves. The PACF plot shows no clear change.

## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.
#note from class: it's just because the detrending wasn't performed very well. in reality, detrend it first, and use the detrened one!

### Q6
Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.

#note from class: look at the overall p-value rather than the one for each p-values: it means there is no significant difference from the intercept, not meaning the seasonal mean model doesn't work well
```{r}
dummies <- seasonaldummy(ts_energy_data[,1])

#Biomass energy production
seas_means_model_biomass <- lm(ts_energy_data[,1]~dummies)
summary(seas_means_model_biomass)
beta_int_biomass=seas_means_model_biomass$coefficients[1]
beta_coeff_biomass=seas_means_model_biomass$coefficients[2:12]

#Renewable energy production
seas_means_model_renewable <- lm(ts_energy_data[,2]~dummies)
summary(seas_means_model_renewable)
beta_int_renewable=seas_means_model_renewable$coefficients[1]
beta_coeff_renewable=seas_means_model_renewable$coefficients[2:12]

#hydroelectric power consumption
seas_means_model_hydro <- lm(ts_energy_data[,3]~dummies)
summary(seas_means_model_hydro)
beta_int_hydro=seas_means_model_hydro$coefficients[1]
beta_coeff_hydro=seas_means_model_hydro$coefficients[2:12]

```
According to the regression output, the p-values in the first two seasonal means models are all greater than 0.05. This indicates that the seasonal means model does not fit the seasonal components well.
On the other hand, the seasonal means model works well for the time series of hydroelectric power consumption. Most of the p-values are less than 0.05. The intercept shows the seasonal mean for the month of December, while the other coefficients show the relative change to the seasonal mean of December.

### Q7
Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

Answer:
```{r}
#plot biomass energy production with its seasonal component
biomass_seas_comp=array(0,nobs)
for(i in 1:nobs){
  biomass_seas_comp[i]=(beta_int_biomass+beta_coeff_biomass%*%dummies[i,])
}
deseason_biomass <- energy_data[,1] - biomass_seas_comp
ts_deseason_biomass <- ts(deseason_biomass)
print(
  autoplot(ts_deseason_biomass)+
    labs(y='Trillion btu', title = datanames[1])+
    geom_line(aes(y=ts_energy_data[,1]),color='brown')
)


#same for renewable and hydro
renewable_seas_comp=array(0,nobs)
for(i in 1:nobs){
  renewable_seas_comp[i]=(beta_int_renewable+beta_coeff_renewable%*%dummies[i,])
}
deseason_renewable <- energy_data[,2] - renewable_seas_comp
ts_deseason_renewable <- ts(deseason_renewable)
print(
  autoplot(ts_deseason_renewable)+
    labs(y='Trillion btu', title = datanames[2])+
    geom_line(aes(y=ts_energy_data[,2]),color='green')
)

hydro_seas_comp=array(0,nobs)
for(i in 1:nobs){
  hydro_seas_comp[i]=(beta_int_hydro+beta_coeff_hydro%*%dummies[i,])
}
deseason_hydro <- energy_data[,3] - hydro_seas_comp
ts_deseason_hydro <- ts(deseason_hydro)
print(
  autoplot(ts_deseason_hydro)+
    labs(y='Trillion btu', title = datanames[3])+
    geom_line(aes(y=ts_energy_data[,3]),color='blue')
)


```
Yes. Not only do the values decrease after deseasoning, but the random fluctuations in each line seem to be more noticeable.

### Q8
Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?
#what do we ultimately want is a fast decaying ACF&PACF plot, meaning that we donnot need to take much lags into consideration

```{r}
#Biomass energy production
par(mfrow=c(1,2))
Acf(ts_energy_data[,1],lag.max=40,main=datanames[1],cex.main=0.3,ylim=c(-1,1))
Acf(ts_deseason_biomass,lag.max = 40, main=paste(datanames[1],"(deseasoned)"),ylim=c(-1,1),cex.main=0.3, )
Pacf(ts_energy_data[,1],lag.max=40,main=datanames[1],cex.main=0.3,ylim=c(-1,1))
Pacf(ts_deseason_biomass,lag.max = 40, main=paste(datanames[1],"(deseasoned)"),ylim=c(-1,1),cex.main=0.3)

#Renewable energy production
par(mfrow=c(1,2))
Acf(ts_energy_data[,2],lag.max=40,main=datanames[2],cex.main=0.3,ylim=c(-1,1))
Acf(ts_deseason_renewable,lag.max = 40, main=paste(datanames[2],"(deseasoned)"),ylim=c(-1,1),cex.main=0.3, )
Pacf(ts_energy_data[,2],lag.max=40,main=datanames[2],cex.main=0.3,ylim=c(-1,1))
Pacf(ts_deseason_renewable,lag.max = 40, main=paste(datanames[2],"(deseasoned)"),ylim=c(-1,1),cex.main=0.3)

#Hydro energy consumption
par(mfrow=c(1,2))
Acf(ts_energy_data[,3],lag.max=40,main=datanames[3],cex.main=0.3,ylim=c(-1,1))
Acf(ts_deseason_hydro,lag.max = 40, main=paste(datanames[3],"(deseasoned)"),ylim=c(-1,1),cex.main=0.3, )
Pacf(ts_energy_data[,3],lag.max=40,main=datanames[3],cex.main=0.3,ylim=c(-1,1))
Pacf(ts_deseason_hydro,lag.max = 40, main=paste(datanames[3],"(deseasoned)"),ylim=c(-1,1),cex.main=0.3)

```

Yes, all plots changed. Any wave-like patterns in original ACF and PACF plots no longer exist. In addition, the PACF values decay much faster after deseasoning. There are very few values remaining above significant levels, most of which are located at lags close to 0.